---
title: "Analysis"
author: "Allen Hulbert, Florian Hartig"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    self-contained-math: true
editor: visual
editor_options: 
  chunk_output_type: console
params:
  seed: NULL    # default, will be overridden by render()
---

```{r, echo=FALSE}
cacheData = F
```

Required packages

```{r, message=FALSE}
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ranger)
library(stringr)
library(heatmaply)
library(usethis)
library(fields)
```

Script settings

```{r}
set.seed(12345)
bootstrap = params$seed
chunkEval = is.null(bootstrap)
resultsToSave = list()
rangerTreads = 4
```

# Data preparation

## Calculate tree metrics and align with model processes

```{r}
source('code/analysis_functions.R')
source('code/plot_functions.R')

#metricsForManyTrees(list.files("trees/uniform_sampling_experiment"),
#                    treedir = "trees/uniform_sampling_experiment",
#                    fileOut = "derived_tree_data/USE_treeStats.txt")


# Parameter key across simulations; 
# url <- 'https://docs.google.com/spreadsheets/d/1pcUuINauW11cE5OpHVQf_ZuzHzhm2VJkCn7-lSEJXYI/edit#gid=1171496897'
# paramKey <- gsheet2tbl(url)
# write.csv(paramKey, 'trees/uniform_sampling_experiment/paramKeys/simulation_parameters_key.csv', row.names = F)

# Align parameter values with the relevant experiment, create a dataframe with columns for
# model, model2 (with scenario suffix if appropriate), simID, and columns for the parameter
# names and parameter values associated with 5 processes: env, dis, nic, mut, com

modelList = c('ca', 'fh', 'gen', 'hs', 'pontarp', 've', 'xe', 'la')

for (m in modelList) {
  
  if (!exists("processDF")) {
    processDF = alignParametersWithProcesses(m, paramKeyPath = 'trees/uniform_sampling_experiment/paramKeys')
  } else {
    processDF = rbind(processDF, alignParametersWithProcesses(m, paramKeyPath = 'trees/uniform_sampling_experiment/paramKeys'))
  }
  
}
```

Join tree metrics to the aligned parameter-process dataframe: for analysis, use **processDFmetrics** Script for calculating tree metrics for trees that have not already been analyzed

```{r}
# Read in existing output file
treeOutput = read.table('derived_tree_data/USE_treeStats.txt', sep = '\t', header = T)

# Join process-parameter linkage dataframe to tree output
allSimuData = left_join(processDF, treeOutput, by = c('model', 'simID'))

# convert to factor
allSimuData[sapply(allSimuData, is.character)] <- lapply(allSimuData[sapply(allSimuData, is.character)], 
                                       as.factor)
```

## Selection of the data

-   filter out pweird
-   remove tree_height and crown_age as well as mntd and mpd which are redundant with mean_branch_length_ext and psv, respectively
-   exclude trees with fewer than 3 tips.

PLUS: bootstrap if bootstrap is set

```{r}
# For some reason there are 300 pontarp simulations with identical parameter sets:
pweird = filter(allSimuData, model == "pontarp", env1 > .9, env1 < .95, 
                dis1 > .09, dis1 < .095, com1 > .26, com1 < .27, 
                nic1 > -.039, nic1 < -.021)

if (is.null(bootstrap)){
simuData = allSimuData %>%
  filter(!(model == "pontarp" & simID %in% pweird$simID),
         number_of_lineages >= 3) %>%
  dplyr::select(-tree_height, -crown_age, -mntd, -mpd)  
} else {
  set.seed(bootstrap)
simuData = allSimuData %>%
  filter(!(model == "pontarp" & simID %in% pweird$simID),
         number_of_lineages >= 3) %>%
  dplyr::select(-tree_height, -crown_age, -mntd, -mpd) %>%
  group_by(model) %>%
  # Bootstrapping here, resampling the simulations with replacement,
  # prop is the proportion of group to resample (i.e. entire group)
  slice_sample(prop = 1, replace = TRUE) %>% 
  ungroup()
}
simuData = as.data.frame(simuData)
```

## Load simulation / bootstrap results

```{r}
bootResults = list()

files = list.files(path = "./bootstrap/", pattern = "*.Rdata") 

for (i in 1:length(files)) {
  load(file = paste("./bootstrap/", files[i], sep = ""))
  bootResults[[i]] = resultsToSave
}

nBoot = length(files) - 1


```

### Case Study

# Correlation of metrics

The following code illustrates the pairwise correlations among tree shape metrics, first for all metrics, and second for just the set of metrics that are not strongly (r \> 0.9) correlated with species richness.

```{r}
# Visualize correlation matrix of tree metrics
metricCors = cor(simuData[, 24:ncol(simuData)], use = "pairwise.complete.obs")

# Metrics strongly correlated (> 0.9) with number_of_lineages
richnessCorrelatedMetrics = row.names(metricCors)[metricCors[,1] >= 0.9]

# all 50 tree metrics
statisticsIndices1 = which(names(simuData) %in% row.names(metricCors))
# Only the tree metrics not strongly correlated with richness
statisticsIndices2 = which(names(simuData) %in% row.names(metricCors) & 
                             !names(simuData) %in% richnessCorrelatedMetrics)

# Indices for the 5 simulation processes (env, dis, mut, nic, com)
predictorsIndices = c(14,16,18,20,22)
predictors = colnames(simuData)[c(14,16,18,20,22)]
numModelsPerPredictor = c(5, 7, 4, 7, 6)

# 8 models (accounting for abbreviation changes post-experiment)
modelAbbrevs = data.frame(original = c('ca', 'fh', 'gen', 'hs', 'la', 'pontarp', 've', 'xe'), 
                    new = c('ca', 'fh', 'ge', 'hs', 'am', 'pw', 've', 'xe'),
                    env = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE),
                    dis = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE),
                    nic = c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE),
                    mut = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE),
                    com = c(TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, FALSE, TRUE))

models = modelAbbrevs$original
```

## Plot correlation of all metrics (Fig. S1)

```{r}
corrs = cor(simuData[, statisticsIndices1], use = "pairwise.complete.obs", )

metricNames = colnames(simuData)[statisticsIndices1]

colnames(corrs) = metricNames
rownames(corrs) = metricNames

resultsToSave$correlation$corrs = corrs
```

Get clustering order of metric names for Figure 2 based on absolute value of correlations

```{r}

dissim = 1 - abs(corrs)
distances = as.dist(dissim)
hc = hclust(distances)
dend = as.dendrogram(hc)
dend.order = order.dendrogram(dend)
```

```{r, fig.height= 10, fig.width=10}
par(mar = c(8, 8, 10, 8))
heatmaply(corrs, colors = cool_warm)
```

The following code for automatic printout doesnt on my computer work unfortunately

```{r, eval = F}
heatmaply(corrs, colors = cool_warm, file = 'figures/Fig.S1-MetricsCorrelation.pdf')
```

## Plot correlation of metrics not strongly correlated with richness

```{r}
corrs2 = cor(simuData[, statisticsIndices2], use = "pairwise.complete.obs", )

metricNames2 = colnames(simuData)[statisticsIndices2]

colnames(corrs2) = metricNames2
rownames(corrs2) = metricNames2
```

```{r, fig.height= 10, fig.width=10}
par(mar = c(8, 8, 10, 8))
heatmaply(corrs2, colors = cool_warm)
```

# Consistency analysis

## Reading in calculations

```{r}
# loading original results
R2MatFull = bootResults[[1]]$consistency$R2MatFull
pMatFull = bootResults[[1]]$consistency$pMatFull
agreement = bootResults[[1]]$consistency$agreement

#creating bootstrap matrices
R2MatFullBoot = pMatFullBoot = array(dim = c(nBoot, dim(R2MatFull)))

# writing results in matrices
for (i in 2:length(files)) {
  R2MatFullBoot[i-1, , ,] = bootResults[[i]]$consistency$R2MatFull
  pMatFullBoot[i-1, , ,] = bootResults[[i]]$consistency$pMatFull  
} 

R2MatFullBootSD <- apply(R2MatFullBoot, c(2,3,4), sd )
pMatFullBootSD <- apply(pMatFullBoot, c(2,3,4), sd )

dimnames(R2MatFullBootSD) = dimnames(pMatFullBootSD) = dimnames(R2MatFull)
```

## Fig. 1 Consistency

Correlation plot highlighting metrics with strong correlation agreement

Two versions, one with all metrics, and one only including one metric per correlated cluster.

Eg, mntd is strongly correlated with mean_branch_length_ext and mean_branch_length_ext, so only one is chosen to be included in the plot based on which is likely to be more familiar or intuitive to readers.

```{r}
env.metrics = rownames(agreement)[agreement[,1] >= 5]
dis.metrics = rownames(agreement)[agreement[,2] >= 7]
nic.metrics = rownames(agreement)[agreement[,3] >= 4]
mut.metrics = rownames(agreement)[agreement[,4] >= 7]
com.metrics = rownames(agreement)[agreement[,5] >= 6]

# Correlation plot with ALL tree metrics for supplement

# Tidier labels
alt.x.labels = dimnames(R2MatFull)[[1]]

# Modifying a few model abbreviations here (e.g. am and pw)
dimnames(R2MatFull)[[3]] = c("ca", "fh", "ge", "hs", "am", "pw", "ve", "xe")
resultsToSave$consistency$R2MatFull = R2MatFull
resultsToSave$consistency$pMatFull = pMatFull
```

boxes = F so that they can be plotted manually in larger blocks Run `plot(dend, edgePar = list(lwd=2))` and copy and paste dendrogram which must be both flipped horizontally and then rotated 90 deg to add to the right side of the all_metrics_correlations.pdf plot. Best if copied from window with wide x short dimensions.

```{r}
pdf('figures/Fig.1-Consistency.pdf', height = 15, width = 13)
par(mfrow = c(1,1), mar = c(4,15,3,3), oma = c(3, 0, 2, 0))

modelMetricDivergent(R2MatFull, pMatFull, x.labels = alt.x.labels, y.labels = c("Env. filtering", "Dispersal", "Niche conserv.", "Speciation", "Competition"),
             cex.metrics = 1.4, cex.models = 1.3, cex.processes = 1.4,
             boxes = T, 
             boxProcesses = c(rep('env', length(env.metrics)),
                                         rep('dis', length(dis.metrics)),
                                         rep('mut', length(mut.metrics))), 
             boxMetrics = c(env.metrics, dis.metrics, mut.metrics), 
             lwd = 5, srt = 90, cex.asterisk = 2, metric.order = dend.order)

mtext("Models", 1, outer = TRUE, cex = 2.5, at = .53)
#mtext("Metrics", 2, cex = 2.5, outer = TRUE, line = -3)
mtext("Processes", 3, cex = 2.5, outer = TRUE, at = 0.53)
dev.off()
```

## Fig. S2 - Bootstrap

```{r}
pdf('figures/Fig.S2-ConsistencyBootSD.pdf', height = 15, width = 13)
par(mfrow = c(1,1), mar = c(4,15,3,3), oma = c(3, 0, 2, 0))

modelMetricContinous(R2MatFullBootSD,
                   cex.metrics = 1.4, cex.models = 1.3, cex.processes = 1.4,
              lwd = 5, srt = 90)
mtext("Models", 1, outer = TRUE, cex = 2.5, at = .53)
#mtext("Metrics", 2, cex = 2.5, outer = TRUE, line = -3)
mtext("Processes", 3, cex = 2.5, outer = TRUE, at = 0.53)
dev.off()
```

# Model inversion

Because simulation models differ substantially in the size of trees they typically produce, we run the random forests and conduct model inversions using only the tree metrics that are not strongly correlated with richness.

## Fig. S3 - Variable Importance

Reading in calculations

```{r}
RFimportance = bootResults[[1]]$predictability$importance
```

```{r}
pdf('figures/Fig.S3-VariableImportance.pdf', height = 15, width = 13)
par(mfrow = c(1,1), mar = c(4,15,3,3), oma = c(3, 0, 2, 0))

modelMetricContinous(RFimportance,
                   cex.metrics = 1.4, cex.models = 1.3, cex.processes = 1.4,
              lwd = 5, srt = 90)
mtext("Models", 1, outer = TRUE, cex = 2.5, at = .53)
#mtext("Metrics", 2, cex = 2.5, outer = TRUE, line = -3)
mtext("Processes", 3, cex = 2.5, outer = TRUE, at = 0.53)
dev.off()
```

## Fig. 2 - predictability within a model

Reading in calculations

```{r}
predictabilityR2CV = bootResults[[1]]$predictability$predictabilityR2CV
```

The internally calculated R2 values of a RF can sometimes be misleading. That's why we additionally calculated R2 values from a 5-fold CV in the calculation loop above. Results are very similar.

```{r}
modelColors = data.frame(model = models,
                         newLabel = c('ca', 'fh', 'ge', 'hs', 'am', 'pw', 've', 'xe'),
                         color = c(rgb(0,0,0),
                                   rgb(230/255, 159/255, 0),
                                   rgb(86/255, 180/255, 233/255),
                                   rgb(0, 158/255, 115/255),
                                   rgb(240/255, 228/255, 66/255),
                                   rgb(0, 114/255, 178/255),
                                   rgb(213/255, 94/255, 0),
                                   rgb(204/255, 121/255, 167/255)))
```

```{r}
pdf('figures/Fig.2-predictabilityWithin.pdf', height = 8, width = 14)

par(mar = c(4, 6, 0, 0), mgp = c(4, 1, 0), mfrow = c(1,1))

# Squaring the correlation values of predictabilityR2CV to get R2's
x = barplot(t((predictabilityR2CV))^2, beside = T, las = 1, horiz = F, ylim = c(0 ,1.1), ylab = "Within model predictability", xaxt = "n", cex.lab = 1.8, cex.axis = 1.6, col = modelColors$color)

# Add 'x' where process is missing from model
missing.x = x[is.na(t(predictabilityR2CV))]
text(x = missing.x, y = rep(0.02, length(missing.x)), 'X', cex = 1.5)

axis(1, at = c(5, 14, 23, 32, 41), labels = c("Env.\nfiltering", "Dispersal", "Niche\nconserv.", "Speciation", "Competition"), cex.axis = 1.6, line = 1.3, tick = F)

axis(1, at = c(0, 9.5, 18.5, 27.5, 36.5, 45.5), tick = T, tck = -0.04, labels = F)

# Legend
text(6, 1.01, "Models", cex = 2)
legend(0, 1, legend = modelColors$newLabel, fill = modelColors$color, ncol = 2, bty = 'n', cex = 1.8)

dev.off()
```

## Fig. 3 - predictability between models

As a next step, we calculate show between models, i.e. we use the random forests tuned for one model to predict to the next model. Cross-model predictability is measured by Kendall's rank correlation coefficient. For the diagonal (within-model predictability), we use the CV predictability calculated in the previous step.

Reading in calculations

```{r}
speciationCP = bootResults[[1]]$crossPredictability$speciationCP
dispersalCP = bootResults[[1]]$crossPredictability$dispersalCP

#creating bootstrap matrices
speciationCPBoot = dispersalCPBoot = array(dim = c(nBoot, dim(speciationCP)))

# writing results in matrices
for (i in 2:length(files)) {
  speciationCPBoot[i-1, , ] = bootResults[[i]]$crossPredictability$speciationCP
  dispersalCPBoot[i-1, , ] = bootResults[[i]]$crossPredictability$dispersalCP  
} 

speciationCPBootSD <- apply(speciationCPBoot, c(2,3), sd )
dispersalCPBootSD <- apply(dispersalCPBoot, c(2,3), sd )

dimnames(speciationCPBootSD) = dimnames(speciationCP)
dimnames(dispersalCPBootSD) = dimnames(dispersalCP)
```

2-panel figure showing cross-predictability for a) dispersal and b) speciation This takes several minutes.

```{r}

pdf('figures/Fig.3-predictabilityBetween.pdf', height = 7, width = 14)

par(mfrow = c(1, 2), mar = c(6,7,4,4), mgp = c(3, 1, 0), oma = c(0, 0, 0, 1))

crossPredictabilityPlot(speciationCP, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5,
                    cexLabSub = 1.5,
                    mainText = "Speciation")
mtext("a)", 2, at= 1.2, las = 1, cex = 2.5, line = 2)

crossPredictabilityPlot(dispersalCP, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5,
                    cexLabSub = 1.5, mainText = "Dispersal")
mtext("b)", 2, at = 1.2, las = 1, cex = 2.5, line = 2)

dev.off()
```

## Fig S4 - Bootstrap uncertainty

```{r}
pdf('figures/Fig.S4-predictabilityBetween-Boot.pdf', height = 7, width = 14)

par(mfrow = c(1, 2), mar = c(6,7,4,4), mgp = c(3, 1, 0), oma = c(0, 0, 0, 1))

crossPredictabilityPlot(speciationCPBootSD, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5,
                    cexLabSub = 1.5,
                    range = c(0,1),
                    xCol = rev(c("darkblue","blue","lightblue", "white")),
                    mainText = "Speciation")
mtext("a)", 2, at= 1.2, las = 1, cex = 2.5, line = 2)


crossPredictabilityPlot(dispersalCPBootSD, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5,
                    cexLabSub = 1.5,
                    range = c(0,1),
                    xCol = rev(c("darkblue","blue","lightblue", "white")),
                    mainText = "Speciation")
mtext("a)", 2, at= 1.2, las = 1, cex = 2.5, line = 2)

dev.off()
```

## Fig S5

2-panel figure showing cross-predictability for a) environmental filtering and b) competition. This takes several minutes.

```{r}
envCP = bootResults[[1]]$crossPredictability$envCP
comCP = bootResults[[1]]$crossPredictability$comCP
nicCP = bootResults[[1]]$crossPredictability$nicCP
```

```{r}
pdf('figures/Fig.S3-predictabilityBetweenEnvComp.pdf', height = 14, width = 14)

par(mfrow = c(2, 2), mar = c(6,7,4,4), mgp = c(3, 1, 0), oma = c(0, 0, 0, 1))

crossPredictabilityPlot(envCP, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5, cexLabSub = 1.5,
                    mainText = "Envir. Filtering")
mtext("a)", 2, at= 1.2, las = 1, cex = 2.5, line = 2)

crossPredictabilityPlot(comCP, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5, cexLabSub = 1.5,
                    mainText = "Competition")
mtext("b)", 2, at = 1.2, las = 1, cex = 2.5, line = 2)

crossPredictabilityPlot(nicCP, cexLab = 2,
                    cexAxis = 1.8, cexValues = 1.3, cexMain = 2.5, cexLabSub = 1.5,
                    mainText = "Niche conservatism")
mtext("c)", 2, at= 1.2, las = 1, cex = 2.5, line = 2)

dev.off()
```

# Case study

General idea: Fit random forests for each simulation model on the simulated data to predict model parameter values from simulation-derived tree shape metrics. Then, use empirical tree shape metrics to infer process parameter values based on the fitted RF for each of these models.

Note that in this simulation, RF that were fit excluded all tree shape metrics correlated to tree size.

## Fig. 4

Read in data

```{r}
scaledPredictions2 = bootResults[[1]]$casestudy$scaledPredictions2

empiricalMetrics = bootResults[[1]]$casestudy$empiricalMetrics

#creating bootstrap matrices

tmp = array(dim = c(nBoot, nrow(scaledPredictions2)))

# writing results in matrices
for (i in 2:length(files)) {
  tmp[i-1, ] = bootResults[[i]]$casestudy$scaledPredictions2$scaledValue
} 

scaledPredictions2$scaledValueSD = apply(tmp, 2, sd)
```

Function for plotting inferred process strengths of empirical sister clades, as well as actual tree metric values for those clades. bottom panel illustrates relative magnitude of two metrics which were identified as the metrics with consistent responses to specified process

Note: needs to be kept here because the function accesses global variables

```{r}
empiricalModelInferencePlot = function(process, # env1, dis1, mut1, nic1, com1
                                       metric1, metric2=NULL, 
                                       met1label, met2label=NULL, 
                                       pairOrder = c(1, 2, 5, 4, 6, 8),
                                       lwd = 3,
                                       cex.axis = 1,
                                       cex.clade = 1.1,
                                       cex.metrics = .8,
                                       cex.models = 1,
                                       cex.process = 1.1,
                                       metric.offset = .22,
                                       cex.quantiles = .9,
                                       cex.panel = 2,
                                       cex.legend = 1.5,
                                       cex.legendpts = 2,
                                       line.process = 2,
                                       line.quantile = 2.1,
                                       line.clade = 2.5, 
                                       line.metrics = 0.6,
                                       cex.axis.quantiles = 1.5,
                                       legend = FALSE,
                                       model.abbrev = TRUE,
                                       panel.margins = c(6, 5, 1, 1),
                                       outer.margins = c(0, 3, 0, 0),
                                       bar.space = .1,
                                       errorBars = FALSE,
                                       lwdErrorBars = 2) {
  
  par(mfrow = c(2, 3), mar = panel.margins, oma = outer.margins)
  
  panelLetters = letters[1:6]
  panelIndex = 0
  
  processes = c('env1', 'dis1', 'mut1', 'nic1', 'com1')
  processLabels = c('Env. Filtering', 'Dispersal', 'Speciation', 'Niche conserv.',
                    'Competition')
  
  for (p in pairOrder) {
    
    panelIndex = panelIndex + 1
    
    tmp = filter(scaledPredictions2, pair == p, predictor == process) %>% 
      left_join(empiricalMetrics[, c('tree2', 'cladeName')], by = c('tree' = 'tree2'))
    
    plot(c(0.5,2.5), range(tmp$scaledValue), type = 'n', xaxt = 'n', xlab = "",
         ylab = "", las = 1, ylim = c(-.3, 1), yaxt = 'n')
    
    mtext(tmp$cladeName[1:2], 1, at = 1:2, line = line.clade, cex = cex.clade, padj = 0.5)
    offset = metric.offset
    
    axis(2, at = seq(0.1, 1, by = 0.1), las = 1, cex.axis = cex.axis)
    
    abline(h = 0.1)
    
    for (m in unique(tmp$model)) {
      
      points(tmp$clade[tmp$model==m], tmp$scaledValue[tmp$model==m], type = 'b',
             col = tmp$color[tmp$model==m], lwd = lwd)
      
      if(errorBars) {
        segments(tmp$clade[tmp$model==m], tmp$scaledValue[tmp$model==m] - 
                   tmp$scaledValueSD[tmp$model==m],
                 tmp$clade[tmp$model==m], tmp$scaledValue[tmp$model==m] + 
                   tmp$scaledValueSD[tmp$model==m],
                 col = tmp$color[tmp$model==m],
                 lwd = lwdErrorBars)
      }
      
      if (model.abbrev) {
        text(2.1, tmp$scaledValue[tmp$model==m & tmp$clade == 2],
             tmp$newLabel[tmp$model==m], adj = 0, cex = cex.models)
      }
    } #end models
    
    # panel label
    text(.58, 0.97, paste0("(", panelLetters[panelIndex], ")"), cex = cex.panel)
    
    # Add barplots showing empirical values of mean.Iprime and VRD for the two sister clades
    par(new = T)
    
    if(!is.null(metric2)) { # if there are 2 metrics to plot
      
      tmpBarData = empiricalMetrics[empiricalMetrics$pair == p, 
                                    c(metric1, metric2)] %>%
        rowwise() %>%
        mutate(quantile.met1 = sum(get(metric1) > simuData[, metric1], 
                                   na.rm = T)/nrow(simuData),
               quantile.met2 = sum(get(metric2) > simuData[, metric2], 
                                   na.rm = T)/nrow(simuData)) %>%
        dplyr::select(quantile.met1, quantile.met2) %>%
        t() %>%
        as.matrix()
      
      barplot(tmpBarData, beside = T, yaxt = 'n', ylim = c(0, 3), border = NA, 
              xlim = c(0.5, 7), xaxt = 'n', space = c(bar.space, 1.4-bar.space))
      axis(2, at = c(0, 0.25, 0.5, 0.75), labels = c(0, 25, 50, 75), las = 1, 
           cex.axis = cex.axis.quantiles)
      abline(h = c(.25, .5, .75), col = 'gray', lty = 'dashed')
      
      mtext(c(met1label, met2label, met1label, met2label), 1, at = c(1.1 - offset, 1.05 + offset, 2.05 - offset, 2. + offset), line = line.metrics, cex = cex.metrics)
      
      
    } else { # if there is 1 metric to plot
      
      tmpBarData = empiricalMetrics[empiricalMetrics$pair == p, 
                                    c('simID', metric1)] %>%
        rowwise() %>%
        mutate(quantile.met1 = sum(get(metric1) > simuData[, metric1], 
                                   na.rm = T)/nrow(simuData)) %>%
        dplyr::select(quantile.met1) %>%
        t() %>%
        as.matrix()
      
      barplot(tmpBarData, yaxt = 'n', ylim = c(0, 3), border = NA, 
              xlim = c(0, 1.3), xaxt = 'n', width = 0.5)
      axis(2, at = c(0, 0.25, 0.5, 0.75), labels = c(0, 25, 50, 75), las = 1, 
           cex.axis = cex.axis.quantiles)
      abline(h = c(.25, .5, .75), col = 'gray', lty = 'dashed')
      
      #mtext(c(met1label, met2label, met1label, met2label), 1, at = c(1.1 - offset, 1.05 + offset, 2.05 - offset, 2. + offset), line = line.metrics, cex = cex.metrics)
      
    }
    
    
    if (panelIndex %in% c(1, 4)) {
      
      yaxis.string = ifelse(is.null(metric2),
                            paste(met1label,"\nquantile"),
                            "Tree Metric\nquantile")
      
      mtext(yaxis.string, 2, cex = cex.quantiles, line = line.quantile, at = 0.37)
      mtext(paste("Inferred", processLabels[processes == process]), 2, cex = cex.process, line = line.process, at = 2)
    }
    
    #if (legend & panelIndex %in% c(1,6)) {
    legend("topright", legend = tmp$newLabel[tmp$clade == 1], 
           col = tmp$color[tmp$clade == 1], pch = 15, #bty = 'n', 
           cex = cex.legend, pt.cex = cex.legendpts)
    #}
  } # end pairOrder
  
} # end function
```

Plot of model inversion for dispersal across 6 empirical sister clade pairs -bottom panel illustrates relative magnitude of mntd and laplac_spectrum_p which were identified as the tree metrics with consistent responses to speciation

```{r}

pdf('figures/Fig.4.pdf', height = 14, width = 10)

empiricalModelInferencePlot('mut1', metric1='mean_branch_length', 
                            metric2=NULL,
                              met1label = 'MBL', 
                             model.abbrev = F, legend = TRUE, cex.legend = 2, 
                             cex.legendpts = 2.5, cex.metrics = 1.5, 
                             metric.offset = .24, cex.clade = 1.4, 
                             cex.quantiles = 1.5, cex.panel = 3, 
                             cex.axis = 1.75, bar.space = .35, 
                             cex.process = 1.7, line.process = 5, 
                            line.quantile = 4, line.clade = 1.5, 
                            line.metrics = .9, cex.axis.quantiles = 1.75, 
                            lwd = 4, outer.margins = c(0, 4, 0, 0))

dev.off()

#empiricalModelInferencePlot('mut1', metric1='mean_branch_length', 
#                            metric2='laplac_spectrum_p',
#                              met1label = 'MBL', met2label = 'LSP', 
#                             model.abbrev = F, legend = TRUE, cex.legend = 2, 
#                             cex.legendpts = 2.5, cex.metrics = 1.5, 
#                             metric.offset = .24, cex.clade = 1.4, 
#                             cex.quantiles = 1.5, cex.panel = 3, 
#                             cex.axis = 1.75, bar.space = .35, 
#                             cex.process = 1.7, line.process = 5, 
#                            line.quantile = 4, line.clade = 3, 
#                            line.metrics = .9, cex.axis.quantiles = 1.75, 
#                            lwd = 4, outer.margins = c(0, 4, 0, 0))



```

Plot of model inversion for dispersal across 6 empirical sister clade pairs bottom panel illustrates relative magnitude of average_leaf_depth and j_one which were identified as the tree metrics with consistent responses to dispersal

```{r}
empiricalModelInferencePlot('dis1', metric1='average_leaf_depth', metric2='j_one',
                              met1label = 'ALD', met2label = 'J_1', 
                             model.abbrev = F, legend = TRUE, cex.legend = 2, 
                             cex.legendpts = 2.5, cex.metrics = 1.5, 
                             metric.offset = .24, cex.clade = 1.4, 
                             cex.quantiles = 1.5, cex.panel = 3, 
                             cex.axis = 1.75, bar.space = .35, 
                             cex.process = 1.7, line.process = 5, line.quantile = 4, line.clade = 3, line.metrics = .9, cex.axis.quantiles = 1.75, lwd = 4, outer.margins = c(0, 4, 0, 0))
```

## Supplemental Figures

```{r}
# Switching out 2 model abbreviations
simuData$model3 = simuData$model
simuData$model3 = gsub("pontarp", "pw", simuData$model3)
simuData$model3 = gsub("la", "am", simuData$model3)
simuData$model3 = gsub("gen", "ge", simuData$model3)

pdf('figures/Fig.SX-metric_boxplots_across_models.pdf', height = 12, width = 10)
par(mfrow = c(7, 5), mar = c(2.5, 2, 2, 1), mgp = c(2.5, .6, 0))

bp1 = boxplot(log10(simuData$number_of_lineages) ~ simuData$model3, main = 'log # of lineages', col = modelColors$color[order(modelColors$newLabel)], xaxt = 'n', tck = -0.01, xlab = '', ylab = '', lwd = .1)
axis(1, at = 1:8, labels = bp1$names, las = 2, cex.axis = 1.4, tck = -0.01)

for (i in statisticsIndices2) { 
  bp = boxplot(simuData[,i] ~ simuData$model3, main = names(simuData)[i], col = modelColors$color[order(modelColors$newLabel)], xaxt = 'n', tck = -0.01, xlab = '', lwd = .1)
  axis(1, at = 1:8, labels = bp$names, las = 2, cex.axis = 1.4, tck = -0.01)
}

dev.off()  
```

# Reproducibility information and saving results

```{r}
sessionInfo()
```

```{r}
bootstrap
if(is.null(bootstrap)) bootstrap = 0
str(resultsToSave)
save(resultsToSave, file = paste("bootstrap/results-", bootstrap, ".Rdata", sep = ""))
```
